Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-11-14 manuscript/manuscript.Rmd
Responding to Rachel's comments

Mesoscale eddies significantly alter open ocean environments such as those found in the subtropical gyres that cover a large fraction of the global ocean. Previous studies have explored eddy effects on biogeochemistry and microbial community composition but not on the molecular composition of particulate organic matter. This study reports the absolute concentration of 67 metabolites and relative abundances for 640 molecular features to understand how mesoscale eddies impact the metabolome of the North Pacific Subtropical Gyre during two cruises in 2017 and 2018. We find that many metabolites track biomass trends, but metabolites like isethionic acid, homarine, and trigonelline linked to eukaryotic phytoplankton were enriched at the deep chlorophyll maximum of the cyclonic features, while degradation products such as arsenobetaine were enriched in anticyclones. In every analysis, metabolites with the strongest responses were detected via untargeted mass spectrometry, indicating that the molecules most sensitive to environmental perturbation were not among the characterized metabolome. By analyzing depth variability (accounting for 20-40% of metabolomic variability across ~150 meters) and the vertical displacement of isopycnal surfaces (explaining 10-20% of variability across a sea level anomaly range of 40 centimeters and a spatial distance of 300 kilometers), this analysis constrains the importance of mesoscale eddies in shaping the chemical composition of particulate matter in the largest biomes on the planet.

@@

The uplift of deep, nutrient-rich seawater into the euphotic zone alters microbial communities [@Rii2022]. Measurements of chlorophyll in cyclones reveal a shallower and more intense maximum [@Barone2022; @Cornec2021] as a result of eukaryotic phytoplankton thriving in the higher nutrient concentrations while cyanobacterial biomass is reduced [@Hawco2021]. This growth of large eukaryotes corresponds to a net increase in biomass and primary productivity [@Benitez-Nelson2007], especially near the deep chlorophyll maximum [@Barone2022]. In contrast, anticyclones produce conditions favorable to nitrogen fixation by accumulating diazotrophs such as *Crocosphaera* and *Trichodesmium* [@Fong2008; @Olson2015; @Dugenne2020; @Dugenne2023].

This shift in community structure and productivity should result in a corresponding shift in the composition of the particulate organic matter produced. Eukaryotic organisms have distinct chemical fingerprints from those of cyanobacteria and heterotrophic bacteria [@Heal2021; @Durham2022] and the relative contribution of these different taxa to total biomass should reflect their chemical composition. However, recent work from @Harke2021 showed that overall community function may be robust across eddy types at the surface while @Gleich2024 detected significant differences in protistan community composition and metabolic potential down to 250 meters, with heterotrophy-associated protistan transcripts enriched in the cyclone. The chemical composition of organic matter is one metric of community function since community metabolism and interactions are in part mediated through organic molecules, making their measurement useful for determining shifts in community dynamics.

In this work, we directly tested for differences in the chemical composition of particulate matter due to changes in eddy state in the NPSG. We sampled from cyclonic and anticyclonic eddies spatially near each other and used targeted and untargeted mass spectrometry to explore changes in the composition of small, polar molecules. We expected to find that overall metabolite abundance would reflect shifts in biomass over multiple depths. We also expected to see enrichment in the deep chlorophyll maximum of cyclonic features for those metabolites especially abundant in eukaryotic organisms. Finally, we predicted that these patterns would be robust across years and sampling regimes for a reliable way to link these ocean features with the chemical composition of organic matter in the open ocean.

@@

Here, we saw how changes in the density surface driven by physics during this process led to the coupling of physics, chemistry, and biology. The enrichment of eukaryotic organisms at the DCM of the cyclone mirrors other ocean processes where upwelling of deep, nutrient-rich water leads to increases in nitrogen-rich metabolites representative of a shift towards organisms such as diatoms and dinoflagellates that undergo bloom dynamics [@Durham2022; @Kulisch2024]. @Barone2022 established a "chemical wake" hypothesis in which cyclone formation introduces a transient phytoplankton community with net photosynthetic production that induces export, a conceptual model that can be extended to include many of the metabolites named here, as some metabolites are more bioavailable than others and the "flavor" of this chemical wake will change over time.

>>>>

Mesoscale eddies significantly alter open ocean environments such as those found in the subtropical gyres that cover a large fraction of the global ocean. Previous studies have explored eddy effects on biogeochemistry and microbial community composition but not on the molecular composition of particulate organic matter. This study reports the absolute concentration of 67 metabolites and relative abundances for 640 molecular features, measured using liquid chromatography-mass spectrometry (LC-MS) following both targeted and untargeted approaches. This approach allowed us to better understand how mesoscale eddies impact the metabolome of the North Pacific Subtropical Gyre during two cruises in 2017 and 2018. We find that many metabolites track biomass trends, but metabolites like isethionic acid, homarine, and trigonelline linked to eukaryotic phytoplankton were enriched at the deep chlorophyll maximum of the cyclonic features, while degradation products such as arsenobetaine were enriched in anticyclones. In every analysis, the metabolites with the strongest responses were identified using LC-MS through untargeted metabolomics approaches, highlighting that the molecules most sensitive to environmental perturbation were not among the previously characterized metabolome. By analyzing depth variability (accounting for 20-40% of metabolomic variability across ~150 meters) and the vertical displacement of isopycnal surfaces (explaining 10-20% of variability across a sea level anomaly range of 40 centimeters and a spatial distance of 300 kilometers), this analysis constrains the importance of mesoscale eddies in shaping the chemical composition of particulate matter in the largest biomes on the planet.

@@

The uplift of deep, nutrient-rich seawater into the euphotic zone alters microbial communities [@Rii2008; @Rii2022]. Measurements of chlorophyll in cyclones reveal a shallower and more intense maximum [@Barone2022; @Cornec2021] as a result of eukaryotic phytoplankton thriving in the higher nutrient concentrations while cyanobacterial biomass is reduced [@Hawco2021]. This growth of large eukaryotes corresponds to a net increase in biomass and primary productivity [@Benitez-Nelson2007], especially near the deep chlorophyll maximum [@Barone2022]. In contrast, anticyclones produce conditions favorable to nitrogen fixation by accumulating diazotrophs such as *Crocosphaera* and *Trichodesmium* [@Fong2008; @Olson2015; @Dugenne2020; @Dugenne2023].

This shift in community structure and productivity should result in a corresponding shift in the composition of the particulate organic matter produced. Eukaryotic organisms have distinct chemical fingerprints from those of cyanobacteria and heterotrophic bacteria [@Heal2021; @Durham2022; @Kuhlisch2023] and the relative contribution of these different taxa to total biomass should reflect their chemical composition. However, recent work from @Harke2021 showed that overall community function may be robust across eddy types at the surface while @Gleich2024 detected significant differences in protistan community composition and metabolic potential down to 250 meters, with heterotrophy-associated protistan transcripts enriched in the cyclone. The chemical composition of organic matter is one metric of community function since community metabolism and interactions are in part mediated through organic molecules, making their measurement useful for determining shifts in community dynamics.

In this work, we directly tested for differences in the chemical composition of particulate matter due to changes in eddy state in the NPSG. We sampled from cyclonic and anticyclonic eddies spatially near each other and used liquid chromatography-mass spectrometry (LC-MS) for both targeted and untargeted metabolomics to explore changes in the composition of small, polar molecules. We expected to find that overall metabolite abundance would reflect shifts in biomass over multiple depths. We also expected to see enrichment in the deep chlorophyll maximum of cyclonic features for those metabolites especially abundant in eukaryotic organisms. Finally, we predicted that these patterns would be robust across years and sampling regimes for a reliable way to link these ocean features with the chemical composition of organic matter in the open ocean.

@@

Here, we saw how changes in the density surface driven by physics during this process led to the coupling of physics, chemistry, and biology. The enrichment of eukaryotic organisms at the DCM of the cyclone mirrors other ocean processes where upwelling of deep, nutrient-rich water leads to increases in nitrogen-rich metabolites representative of a shift towards organisms such as diatoms and dinoflagellates that undergo bloom dynamics [@Durham2022; @Kulisch2023]. @Barone2022 established a "chemical wake" hypothesis in which cyclone formation introduces a transient phytoplankton community with net photosynthetic production that induces export, a conceptual model that can be extended to include many of the metabolites named here, as some metabolites are more bioavailable than others and the "flavor" of this chemical wake will change over time.
